\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-31 standardmodel.tex
Highlight model.

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from~\cite{altkap2016,kaposi2017type, kaposi2019gluing} and their formalization, where they formally resort to the concept of \textit{algebra/model} of QIIT, here we only consider inductively interpreting each syntax piece into naive set theory that is also respecting judgemental equalities\footnote{In our appendix, we still interpret into Agda component to be more precise and less error-prone, also we need the inductive facility from Agda.}

The main idea of the proof is that: (1) we use $\denotesS{}$ to map each intrinsic well-typed syntax piece into a mathematical object; (2) we define proof-irrelevant proposition $\models_S$ for each syntactical $\vdash$, and derive compatibility lemma. We will mainly focus on the definition of $\denotesS{}$ and omit the verification of $\models_S$\footnote{Actually our 1 is acting exactly as an interpretation function, and in our (pseudo) Agda formulation we only do 1 because 2 is automatic; here we make 2 explicit to make our formulation more accessible and uniform}, and also we omit the verification of equality (i.e. the judgementally-equal syntax needs to be mapped to the same data for well-definedness).

The main intuition of the constructed consistency model is that: (1) since we have used sigma type to translate the linkage, using the model for sigma type for linkage will work; (2) directly using the inductive facility of (pseudo)-Agda to justify the intro and elim rule of our W-type.

@@

\newcommand{\goodCtxS}[2]{{ {#1} \ \models_S }}
\newcommand{\goodTypeS}[3]{{ {#1} \models_S {#2} }}
\newcommand{\goodTermS}[3]{{ {#1} \models_S {#2} : {#3} }}
\newcommand{\goodSubS}[3]{{ {#1} \models_S {#2} : {#3} }}
\newcommand{\goodSigS}[3]{{ {#1} \models_S {#2} \ \  Sig^{#3} }}
\newcommand{\goodWSigS}[3]{{ {#1} \models_S {#2} \ \ WSig^{#3} }}
\newcommand{\goodSealS}[4]{{ {#1} \models_S {#2} : {#3} \  |\  {#4} }}
\newcommand{\goodInhS}[4]{{ {#1} \models_S {#2} : {#3} \twoheadrightarrow {#4}}}

@@

  \denotesS{\goodCtx{\Gamma}{i}} & \text{ is a Set and we define } \goodCtxS{\Gamma}{i} :\iff \denotesS{\goodCtx{\Gamma}{i}} \text{ is defined} \\
  \denotesS{\goodType{\Gamma}{T}{i}} & \text{ is a family of sets indexed by } \denotesS{\goodCtx{\Gamma}{i}} \\
  & \text{ and we define } \goodTypeS{\Gamma}{T}{i} : \iff \denotesS{\goodType{\Gamma}{T}{i}} \text{ is defined } \\

@@

  \denotesS{\goodSub{\Gamma}{\_}{\Delta}} & \text{ is a Set and we define }  {\goodSubS{\Gamma}{\gamma}{\Delta}} : \iff \denotesS{\gamma} \in \denotesS{\goodSub{\Gamma}{\_}{\Delta}} \\

@@

  \denotesS{\goodTerm{\Gamma}{\_}{T}} & \text{ is a Set (of dependent function) } \\& \text{and we define }  {\goodTermS{\Gamma}{t}{T}} : \iff \denotesS{t} \in \denotesS{\goodTerm{\Gamma}{\_}{T}} \\

@@

  \denotesS{\goodType{\Gamma}{\Sigma A B}{}} &\text{ is a family of sets, defined dependent on } \denotesS{A}, \denotesS{B} \\

@@

  \denotesS{\goodSig{\Gamma}{\_}{n}} & \text{ is a Set  and we define  } {\goodSigS{\Gamma}{\sigma}{n}} :\iff \denotesS{\sigma} \in  \denotesS{\goodSig{\Gamma}{\_}{n}} \\
  \denotesS{\goodType{\Gamma}{\cC \sigma}{}} & \text{ is a family of sets, dependent on and defined inductively on } \denotesS{\sigma} \\   

@@

  \denotesS{\goodType{\Gamma}{\cL \sigma}{i}} & \text{ is a family of sets, dependent on } \denotesS{\sigma}  \\

@@

  \denotesS{\goodTerm{\Gamma}{\cCt \{\sigma\} o}{\cC\sigma}} & \text{ is a dependent function, defined inductively on } \denotesS{\sigma} \\

@@

  &= \denotesS{\goodTerm{\Gamma}{\cCt (p_1\mu \ o)}{...}}, \denotesS{(p_2\mu \ o)["sf" \ f][(id, \cCt (p_1\mu \ o))]} \\

@@

  \denotesS{\goodWSig{\Gamma}{\_}{n}} &\text{ is a Set of list of pairs of type interpretation of length } n \\
  & \text{ and we define } \goodWSigS{\Gamma}{\tau}{n} :\iff \denotesS{\goodWSig{\Gamma}{\tau}{n}} \in \denotesS{\goodWSig{\Gamma}{\_}{n}}\\

>>>>

\newcommand{\goodCtxS}[2]{{ {#1} \ \models_S }}
\newcommand{\goodTypeS}[3]{{ {#1} \models_S {#2} }}
\newcommand{\goodTermS}[3]{{ {#1} \models_S {#2} : {#3} }}
\newcommand{\goodSubS}[3]{{ {#1} \models_S {#2} : {#3} }}
\newcommand{\goodSigS}[3]{{ {#1} \models_S {#2} \ \  \texttt{LSig}^{#3} }}
\newcommand{\goodWSigS}[3]{{ {#1} \models_S {#2} \ \ \texttt{WSig}^{#3} }}
\newcommand{\goodSealS}[4]{{ {#1} \models_S {#2} : {#3} \  |\  {#4} }}
\newcommand{\goodInhS}[4]{{ {#1} \models_S {#2} : {#3} \twoheadrightarrow {#4}}}

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from~\cite{altkap2016,kaposi2017type, kaposi2019gluing} and their formalization, where they formally resort to the concept of \textit{algebra/model} of QIIT. However, in this section, for readability, we only inductively interpret each syntax piece into naive set theory that is also respecting judgemental equalities. Please refer to our pseduo-Agda formulation for a more formal and complete version.

The big picture of the proof is that: (1) we use $\denotesS{}$ to map each intrinsic well-typed syntax piece into a mathematical object; (2) we define proof-irrelevant proposition $\models_S$ for each syntactical $\vdash$, and derive compatibility lemma. The main goal is to interpret $\denotesS{\bot} = \emptyset$ which implies it is impossible $\goodTermS{\cdot}{t}{\bot}$ and thus $\bot$ is syntactically non-derivable. We will mainly focus on the definition of $\denotesS{}$ and omit the verification of $\models_S$ and that $\denotesS{}$ respecting judgemental equality.\footnote{Actually our 1 is acting exactly as an interpretation function, and in our (pseudo) Agda formulation we only do 1 because 2 is automatic; here we make 2 explicit to make our formulation more accessible and uniform}

The main intuition of the consistency model is that: (1) since we have used sigma type to translate the linkage, using the model for sigma type for linkage will work; (2) directly using the inductive facility of (pseudo)-Agda (or naive set theory) to justify the intro and elim rule of our W-type.

@@

\definecolor{new}{HTML}{F5EFE6}

@@

  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod1}(0.05,-0.15)(-0.10,0.30)
  \denotesS{\goodCtx{\Gamma}{i}} & \text{ is a Set and we define } \goodCtxS{\Gamma}{i} :\iff \denotesS{\goodCtx{\Gamma}{i}} \text{ is defined}\tikzmarkend{stmod1} 

@@

  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod2}(0.05,-0.15)(-0.10,0.30) \denotesS{\goodSub{\Gamma}{\_}{\Delta}} & \text{ is a Set and we define }  {\goodSubS{\Gamma}{\gamma}{\Delta}} : \iff \denotesS{\gamma} \in \denotesS{\goodSub{\Gamma}{\_}{\Delta}} \tikzmarkend{stmod2} 

@@

  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod3}(0.05,-0.15)(-0.10,0.30) \denotesS{\goodTerm{\Gamma}{\_}{T}} & \text{ is a Set (of dependent function) } \\& \text{and we define }  {\goodTermS{\Gamma}{t}{T}} : \iff \denotesS{t} \in \denotesS{\goodTerm{\Gamma}{\_}{T}}
  \tikzmarkend{stmod3}

@@

  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod4}(0.05,-0.15)(-0.10,0.30)
  \denotesS{\goodType{\Gamma}{T}{i}} & \text{ is a family of sets indexed by } \denotesS{\goodCtx{\Gamma}{i}}  
  & \text{ and we define } \goodTypeS{\Gamma}{T}{i} : \iff \denotesS{\goodType{\Gamma}{T}{i}} \text{ is defined } \tikzmarkend{stmod4} 

@@

  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod5}(0.05,-0.15)(-0.10,0.30)
  \denotesS{\goodSig{\Gamma}{\_}{n}} & \text{ is a Set  and we define  } {\goodSigS{\Gamma}{\sigma}{n}} :\iff \denotesS{\sigma} \in  \denotesS{\goodSig{\Gamma}{\_}{n}} \tikzmarkend{stmod5} 
  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod6}(0.05,-0.15)(-0.10,0.30)
  \denotesS{\goodType{\Gamma}{\cC \sigma}{}} & \text{ is a family of sets, defined inductively on } \sigma \tikzmarkend{stmod6} \\   

@@

  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod7}(0.05,-0.15)(-0.10,0.30)
  \denotesS{\goodType{\Gamma}{\cL \sigma}{i}} & \text{ is a family of sets, defined inductively on } \sigma \tikzmarkend{stmod7}  \\

@@

  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod8}(0.05,-0.15)(-0.10,0.30)
  \denotesS{\goodTerm{\Gamma}{\cCt \{\sigma\} o}{\cC\sigma}} & \text{ is a dependent function, defined inductively on } {\sigma} 
  \tikzmarkend{stmod8}

@@

  &= \denotesS{{\cCt (p_1\mu \ o)}}, \denotesS{(p_2\mu \ o)["sf" \ f][(id, \cCt (p_1\mu \ o))]} \\

@@

  \tikzmarkin[disable rounded corners=true,set fill color=new,set border color=new]{stmod9}(0.05,-0.15)(-0.10,0.30)
  \denotesS{\goodWSig{\Gamma}{\_}{n}} &\text{ is a set of list of pairs of type interpretation of length } n \\
  & \text{ and we define } \goodWSigS{\Gamma}{\tau}{n} :\iff \denotesS{\goodWSig{\Gamma}{\tau}{n}} \in \denotesS{\goodWSig{\Gamma}{\_}{n}}
  \tikzmarkend{stmod9}
